### 2025-07-13 docs/article/Sections/benchmark.tex
Post MDPI review changes.

The Spring WebFlux core implementation uses Project Reactor to support a
reactive programming model: each method returns a \texttt{Flux<String>} as the
response body, which acts as a publisher that progressively streams the HTML
content to the client. The implementation includes four main approaches to
PSSR\@:
\begin{itemize}
  \item \textbf{Reactive}: The template engine is used in a reactive context,
        where the HTML content rendered using the reactive programming model.
        An example of this approach is the Thymeleaf template engine
        when using the \texttt{ReactiveDataDriverContextVariable} in conjunction with a
        non-blocking Spring \texttt{ViewResolver}.
  \item \textbf{Virtual}: The template engine is used in a non-blocking
        context, where the HTML content is rendered within the context of Virtual
        Threads. This method is used for the template engines that do not traditionally
        support non-blocking I/O, be it either because they use external DSLs and in
        consequence only support the blocking \texttt{Iterable} interface, or
        because they do not support the asynchronous rendering of HTML content.
  \item \textbf{Suspendable}: The template engine is used in a suspendable
        context, where the HTML content is rendered within the context of a suspending function.
        An example of this approach is the HtmlFlow template engine, which
        supports suspendable templates with the use of the \texttt{Flow} class from the
        Kotlin standard library.
\end{itemize}
% It also includes methods using Kotlin coroutines and other
% asynchronous mechanisms supported by template engines, such as the
% \texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
% using \textit{continuations}. 
The Spring MVC implementation uses handlers based solely on the blocking
interface of the \texttt{Observable} class. To enable PSSR in this context, we
utilize the \texttt{StreamingResponseBody} interface, which allows the
application to write directly to the response \texttt{OutputStream} without
blocking the servlet container thread. According to the Spring documentation,
this class is \textit{a controller method return value type for asynchronous
request processing where the application can write directly to the response
OutputStream without holding up the Servlet container thread.}

@@

However, the Spring MVC implementation does not effectively support PSSR for
these templates, as HTML content is not streamed progressively to the client.
This is because the response is only sent once the content written to the
\texttt{OutputStream} exceeds the output buffer size, which defaults to 8KB\@.
As a result, the client receives the response only after the entire HTML
content is rendered, defeating the purpose of PSSR in this context.

@@

The Quarkus implementation also uses handlers based on the blocking interface
of the \texttt{Observable} class. It implements the \texttt{StreamingOutput}
interface from the JAX-RS specification to enable PSSR, allowing HTML content
to be streamed to the client. While \texttt{StreamingOutput} also uses blocking
I/O, it operates on Vert.x worker threads, which prevents blocking of the event
loop. When Virtual Threads are used, the I/O operations are handled
efficiently, as they are executed in lightweight threads.

>>>>

The Spring WebFlux core implementation uses Project Reactor to support a
reactive programming model: each method returns a \texttt{Flux<String>} as the
response body, which acts as a publisher that progressively streams the HTML
content to the client. The implementation includes four main approaches to
PSSR\@:
\begin{itemize}
      \item \textbf{Reactive}: The template engine is used in a reactive context,
            where the HTML content rendered using the reactive programming model.
            An example of this approach is the Thymeleaf template engine
            when using the \texttt{ReactiveDataDriverContextVariable} in conjunction with a
            non-blocking Spring \texttt{ViewResolver}.
      \item \textbf{Suspendable}: The template engine is used in a suspendable
            context, where the HTML content is rendered within the context of a suspending function.
            An example of this approach is the HtmlFlow template engine, which
            supports suspendable templates with the use of the \texttt{Flow} class from the
            Kotlin standard library.
      \item \textbf{Virtual}: The template engine is used in a non-blocking
            context, where the HTML content is rendered within the context of Virtual
            Threads. This method is used for the template engines that do not traditionally
            support non-blocking I/O, be it either because they use external DSLs and in
            consequence only support the blocking \texttt{Iterable} interface, or
            because they do not support the asynchronous rendering of HTML content.
      \item \textbf{Blocking}: The template engine is used in a blocking
            context, where the HTML content is rendered using the blocking interface of the
            \texttt{Observable} class, within the context of a OS-thread.
\end{itemize}
% It also includes methods using Kotlin coroutines and other
% asynchronous mechanisms supported by template engines, such as the
% \texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
% using \textit{continuations}. 
The Spring MVC implementation uses handlers based solely on the blocking
interface of the \texttt{Observable} class. To enable PSSR in this context, we
utilize the \texttt{StreamingResponseBody} interface, which allows the
application to write directly to the response \texttt{OutputStream} without
blocking the servlet container thread. According to the Spring documentation\footnote{\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}},
this class is \textit{a controller method return value type for asynchronous
request processing where the application can write directly to the response
OutputStream without holding up the Servlet container thread.}

@@

However, the Spring MVC implementation does not effectively support PSSR for
these templates, as HTML content is not streamed progressively to the client.
This limitation occurs because the response is only flushed to the client once
the content written to the \texttt{OutputStream} exceeds the configured output
buffer size, as explained in Section~\ref{sec:web-frameworks}. Consequently,
the client receives the complete response only after the entire HTML content
has been rendered and buffered, which negates the primary benefits of
PSSR—namely, reduced time-to-first-byte and progressive content delivery.
Furthermore, Spring MVC does not provide configuration options for the response
buffer size, preventing developers from reducing it to smaller values that
would enable more frequent flushing and achieve true progressive streaming of
HTML content. This architectural constraint makes Spring MVC unsuitable for
implementing effective PSSR compared to reactive frameworks like Spring
WebFlux.

@@

The Quarkus implementation also uses handlers based on the blocking interface
of the \texttt{Observable} class. It implements the \texttt{StreamingOutput}
interface from the JAX-RS specification to enable PSSR, allowing HTML content
to be streamed to the client. While \texttt{StreamingOutput} also uses blocking
I/O, it operates on Vert.X worker threads, which prevents blocking of the event
loop. When Virtual Threads are used, the I/O operations are handled
efficiently, as they are executed in lightweight threads.

### 2025-07-13 docs/article/Sections/introduction.tex
Post MDPI review changes.

\textit{Progressive Server-Side Rendering (PSSR)} combines benefits from both SSR and CSR
by streaming HTML content progressively. This technique enhances user-perceived performance by
allowing progressive rendering as data becomes available. In this respect,
PSSR is similar to CSR in that the server initially sends a minimal HTML
document to the client and subsequently streams additional HTML fragments.
However, unlike CSR, PSSR retains all rendering responsibilities on the server side,
thereby reducing the load on the client. As a result, the client does
not need to execute JavaScript or make additional requests to retrieve page content.

>>>>

\textit{Progressive Server-Side Rendering (PSSR)} combines benefits from both SSR and CSR
by streaming HTML content progressively. This technique enhances user-perceived performance by
allowing progressive rendering as data becomes available, significantly reducing the
\textit{Time to First Byte (TTFB)} and improving perceived load times compared to traditional
SSR approaches~\cite{wise2024pssr}. In this respect,
PSSR is similar to CSR in that the server initially sends a minimal HTML
document to the client and subsequently streams additional HTML fragments.
However, unlike CSR, PSSR retains all rendering responsibilities on the server side,
thereby reducing the load on the client. As a result, the client does
not need to execute JavaScript or make additional requests to retrieve page content.
The streaming nature of PSSR allows users to see content progressively as it becomes
available, rather than waiting for the complete page to be rendered server-side,
thus providing a more responsive user experience with measurably lower TTFB values.

### 2025-07-13 docs/article/Sections/results.tex
Post MDPI review changes.

All the following benchmarks were conducted on a Github-hosted virtual machine
under GitHub Actions, running Ubuntu 22.04 LTS with a 4-core, 8-thread CPU and
16GB of RAM\@. All tests were conducted on the OpenJDK VM Corretto 21. The JVM
was configured with a minimum heap size of 512MB and a maximum heap size of
16GB\@.

@@

Both the Quarkus and Spring MVC implementations were configured with an output
buffer size of 8KB, despite the Quarkus implementation not enabling PSSR at
this size. This configuration was chosen to ensure consistency across Spring
MVC and Quarkus, as the Spring MVC implementation does not support PSSR for the
tested templates.

Since the obtained results for JMeter and Apache Bench show no significant
differences, only the JMeter results will be presented.

\subsection{Presentations Results}

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
     \caption{Presentation Benchmark Results in Spring WebFlux with JMeter}\label{fig:presentations-webflux-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
     \caption{Presentation Benchmark Results in Spring MVC with JMeter}\label{fig:presentations-springmvc-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
     \caption{Presentation Benchmark Results in Quarkus with JMeter}\label{fig:presentations-quarkus-jmeter}
\end{figure}

@@

\subsection{Stocks Results}

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
     \caption{Stocks Benchmark Results in Spring WebFlux with JMeter}\label{fig:stocks-webflux-jmeter}
\end{figure}

The results in Figure~\ref{fig:stocks-webflux-jmeter} use the same template
engines and approaches as the previous benchmark, but with a more complex data
model: the Stock class, which includes 20 instances and approximately two
times as many data bindings. With this data model, the scalability of the
engines remains largely unchanged; however, throughput is reduced across all
engines. Compared to the Presentation benchmark, the approaches using Virtual
Threads experience a more pronounced decrease in performance relative to the
\textit{Reactive} and \textit{Suspending} approaches, with Jstachio using
Virtual Threads now achieving 1,509 requests per second, compared to 1,750
requests per second achieved by the Thymeleaf implementation using the reactive
View Resolver driver.

It is again important to note that the differences in scalability and
throughput between the \textit{HtmlFlow-Susp}, \textit{Jstachio-Virtual}, and
\textit{Thymeleaf-Rx} approaches may be influenced by the specific template
engines used, rather than the approach itself. When comparing the
\textit{Reactive}, \textit{Suspendable}, and \textit{Virtual} approaches
specifically with HtmlFlow, we found that all three achieve similar
performance: HtmlFlow using a blocking approach with Virtual Threads reaches
3,090 requests per second, while HtmlFlow using a reactive approach achieves
3,026 requests per second.

The overall throughput reduction is expected, as the Stock class contains more
data properties than the \texttt{Presentation} class, adding overhead related
to the data binding process of each template engine.

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
     \caption{Stocks Benchmark Results in Spring MVC with JMeter}\label{fig:stocks-springmvc-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
     \caption{Stocks Benchmark Results in Quarkus with JMeter}\label{fig:stocks-quarkus-jmeter}
\end{figure}

@@

The results of the benchmarks show that non-blocking engines—whether using
reactive programming, Kotlin coroutines, or Java Virtual Threads—are able to
scale effectively, supporting between 32 and 128 concurrent users depending on
the approach and framework. Out of all the tested frameworks, Spring Webflux
showed itself the most effective at enabling PSSR, mostly due to its native
support for Publish and Subscriber interfaces, which allow for HTML content to
be progressively streamed to the client. Quarkus also enabled PSSR effectively,
but it required additional configuration of the \texttt{OutputBuffer} size to
achieve the same results as Spring Webflux. The Spring MVC implementation, on
the other hand, did not enable PSSR for the tested templates.

% Additionally, the results show that approaches using Virtual Threads are able
% to scale as effectively as those using reactive programming or Kotlin
% coroutines, allowing \textit{external} DSLs to be used in a non-blocking
% context, effectively enabling PSSR\@.

>>>>

All the following benchmarks were conducted on a GitHub-hosted virtual machine
under GitHub Actions, running Ubuntu 24.04.2 LTS (Noble Numbat) with an AMD
EPYC 7763 64-Core Processor (x86\_64 architecture) operating at 3.24 GHz,
configured with 2 CPU cores and 2 threads per core, and 7.8GB of available RAM.
The system utilizes a 75GB SSD with ext4 filesystem, achieving 1.5 GB/s write
throughput in basic I/O tests. Network connectivity is provided through a 1500
MTU Ethernet interface. All tests were conducted on OpenJDK 21 (Correto build)
with the G1 garbage collector enabled by default. The following . JVM arguments
were used: \texttt{-Xms512m -Xmx16g}. Each test was executed 5 times per
configuration to ensure the reliability of the results.

All template engines were configured to use UTF-8 encoding and disable
automatic HTML escaping to ensure fair comparison and consistent output across
all template engines. Caching behaviour was left as the default for each
engine, allowing them to optimize template loading and rendering as per their
design. These configurations ensure that all engines operate under equivalent
conditions, with template loading, encoding, and escaping behavior normalized
across implementations.

@@

Both Quarkus and Spring MVC implementations were configured with an 8KB output
buffer size to ensure consistency, despite Quarkus not enabling PSSR at this
size and Spring MVC not supporting PSSR for the tested templates. Testing with
a reduced 512B buffer size showed only a 0.046\% performance difference with
Rocker, indicating negligible impact. Both frameworks use unlimited platform
thread pools, enabling on-demand thread creation up to system limits for
maximum throughput and scalability observation under high concurrent load.

Since the obtained results for JMeter and Apache Bench show no significant
differences, only the JMeter results will be presented. Statistical analysis of
test configurations revealed an average absolute percentage difference of
2.83\% between the two load testing tools. While individual approach
differences ranged from -16.53\% to +14.66\% at the extremes, the consistent
directional bias and small magnitude of differences indicate that both tools
provide comparable performance measurements. Given this statistical equivalence
and the need for brevity, only JMeter results are presented, as they provide
representative performance characteristics across all tested configurations.

\subsection{Scalability results for the \texttt{Presentations} class}

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
     \caption{Throughput (requests per second) scalability results for Spring Webflux with \texttt{Presentations} class}\label{fig:presentations-webflux-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
     \caption{Throughput (requests per second) scalability results for Spring MVC with \texttt{Presentations} class}\label{fig:presentations-springmvc-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
     \caption{Throughput (requests per second) scalability results for Quarkus with \texttt{Presentations} class}\label{fig:presentations-quarkus-jmeter}
\end{figure}

@@

\subsection{Scalability results for the \texttt{Stocks} class}

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
     \caption{Throughput (requests per second) scalability results for Spring Webflux with \texttt{Stocks} class}\label{fig:stocks-webflux-jmeter}
\end{figure}

The results in Figure~\ref{fig:stocks-webflux-jmeter} use the same template
engines and approaches as the previous benchmark, but with a more complex data
model: the Stock class, which includes 20 instances and approximately two times
as many data bindings. With this data model, the scalability of the engines
remains largely unchanged; however, throughput is reduced across all engines.
Compared to the Presentation benchmark, Jstachio using virtual threads
experienced a more pronounced decrease in performance relative to the
\textit{Reactive} and \textit{Suspending} approaches, with Jstachio using
Virtual Threads now achieving 1,509 requests per second, compared to 1,750
requests per second achieved by the Thymeleaf implementation using the reactive
View Resolver driver.

It is again important to note that the differences in scalability and
throughput between the \textit{HtmlFlow-Susp}, \textit{Jstachio-Virtual}, and
\textit{Thymeleaf-Rx} approaches may be influenced by the specific template
engines used, rather than the approach itself. When comparing the
\textit{Reactive}, \textit{Suspendable}, and \textit{Virtual} approaches
specifically with HtmlFlow, we found that all three achieve similar
performance: HtmlFlow using a blocking approach with Virtual Threads reaches
3,090 requests per second, while HtmlFlow using a reactive approach achieves
3,026 requests per second. This indicates that the more pronounced decrease in
performance for Jstachio using Virtual Threads is likely due to the specific
implementation of the template engine, rather than the use of Virtual
Threads itself.

The overall throughput reduction across all engines is expected, as the Stock
class contains more data properties than the \texttt{Presentation} class,
adding overhead related to the data binding process of each template engine.

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
     \caption{Throughput (requests per second) scalability results for Spring MVC with \texttt{Stocks} class}\label{fig:stocks-springmvc-jmeter}
\end{figure}

@@

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
     \caption{Throughput (requests per second) scalability results for Quarkus with \texttt{Stocks} class}\label{fig:stocks-quarkus-jmeter}
\end{figure}

@@

The results of the benchmarks show that non-blocking engines—whether using
reactive programming, Kotlin coroutines, or Java Virtual Threads—are able to
scale effectively, supporting between 32 and 128 concurrent users depending on
the approach and framework. Out of all the tested frameworks, Spring Webflux
showed itself the most effective at enabling PSSR, mostly due to its native
support for publish-subscribe interfaces, allowing for content to be streamed
as data becomes available, instead of when the response buffer is flushed.
Quarkus also enabled PSSR effectively, but it required additional configuration
of the \texttt{OutputBuffer} size to achieve the same results as Spring
Webflux. The Spring MVC implementation, on the other hand, did not enable PSSR
for the tested templates.

However, it is important to acknowledge the limitations of our chosen data
models for generalizability. The tested templates used relatively simple data
structures with limited nesting and straightforward property bindings.
Real-world applications often involve deeply nested data structures, complex
conditional logic, and iterative rendering over thousands of items. Under such
conditions, we anticipate significantly different performance characteristics:
increased memory consumption due to object traversal overhead, higher CPU
utilization for complex evaluations, and altered scalability patterns where
non-blocking approaches may become more advantageous. The performance
degradation observed with our Stock class benchmark—which merely doubled the
number of properties—suggests that enterprise-level template complexity would
result in substantially higher performance degradation. Future work should
investigate these scenarios with more realistic data models to better
understand performance boundaries and optimization strategies for complex PSSR
implementations.

\subsection{Memory Consumption and Resource Utilization Analysis}

This section evaluates the memory and CPU resource characteristics of three
representative concurrency models: Virtual Threads, suspendable coroutines, and
reactive streams. While previous benchmarks demonstrated that Virtual Threads
can offer comparable throughput to non-blocking paradigms, we now assess
whether this simplicity in code comes with higher runtime resource costs.

\textbf{Virtual Threads vs Suspendables}

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/cpu-susp.png}
     \caption{Benchmark CPU utilization profiling with HtmlFlow-Susp in Spring WebFlux}\label{fig:cpu-susp}
\end{figure}

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/cpu-virt.png}
     \caption{Benchmark CPU utilization profiling with HtmlFlow running on Virtual Threads in Spring WebFlux}\label{fig:cpu-virt}
\end{figure}

Figures~\ref{fig:cpu-virt} and~\ref{fig:cpu-susp} show CPU utilization profiles
for the \texttt{HtmlFlow-Virtual} and \texttt{HtmlFlow-Susp} implementations,
respectively. \texttt{HtmlFlow-Virtual} reaches peak CPU usage of around 90\%
during load, with sharp transitions between idle and active phases. In
contrast, \texttt{HtmlFlow-Susp} peaks at approximately 75\% and exhibits
smoother transitions, suggesting more efficient CPU scheduling and lower
contention.

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/gc-susp.png}
     \caption{Benchmark memory utilization profiling with HtmlFlow-Susp in Spring WebFlux}\label{fig:gc-susp}
\end{figure}

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/gc-virt.png}
     \caption{Benchmark memory utilization profiling with HtmlFlow running on Virtual Threads in Spring WebFlux}\label{fig:gc-virt}
\end{figure}

The memory usage disparity is even more pronounced. As shown in
Figure~\ref{fig:gc-virt}, the Virtual Threads implementation gradually climbs
to a peak heap usage of 900MB, with large but infrequent garbage collection
events. Figure~\ref{fig:gc-susp} shows the suspendable version peaks at only
245MB, following a regular sawtooth pattern of small, frequent garbage
collections. Overall, \texttt{HtmlFlow-Virtual} allocates a total of 939MB heap
compared to just 348MB in \texttt{HtmlFlow-Susp}—a 2.7× difference.

\textbf{RxJava Blocking vs Structured Concurrency}

To further assess the efficiency of blocking versus structured concurrency, we
compared the memory behavior of RxJava's \texttt{BlockingObservableIterable}
with \texttt{PublisherAsFlow}. The blocking iterator internally relies on a
lock-based condition variable and an unbounded queue
(\texttt{SpscLinkedArrayQueue}), blocking the thread when no elements are
available. Under high concurrency, this model tends to retain large numbers of
elements in memory, especially when producers outpace consumers. This was
reflected in our profiling, where heap usage for blocking consumption showed
linear growth and irregular GC behavior.

In contrast, \texttt{PublisherAsFlow} employs coroutine-based suspension with
bounded \texttt{Channel} capacity and configurable \texttt{BufferOverflow}
strategies. It requests upstream data based on demand, leveraging structured
concurrency and cooperative cancellation.

It is important to note that our benchmark keeps data items in memory as static
objects, which represents a best-case scenario for memory allocation patterns.
In real-world applications, data typically comes from external sources such as
databases, web services, or file systems, requiring dynamic object creation and
serialization for each request. This would result in significantly higher
memory allocation rates and more frequent garbage collection cycles across all
concurrency models, with the observed differences between Virtual Threads and
suspendable coroutines likely becoming even more pronounced under such
realistic I/O-intensive workloads. While our benchmark demonstrates that
Virtual Threads can achieve comparable throughput to suspendable coroutines, or
other non-blocking paradigms, future work should further explore the impact of
dynamic data sources on memory allocation patterns and garbage collection
behavior to better understand the trade-offs between these concurrency models
in real-world scenarios.

\textbf{Summary of Trade-offs}

While the approaches can achieve similar throughput under load, they differ
significantly in runtime efficiency. Virtual Threads, despite offering a
simplified synchronous style, incur the highest memory and CPU overhead.
structured concurrency delivers superior resource efficiency but
requires familiarity with asynchronous programming constructs. The RxJava
blocking iterator, while easier to adopt in traditional reactive settings,
performs worse in memory-constrained environments due to its unbounded internal
queuing model.

These results suggest that system resource constraints and predictability
requirements should inform the choice of concurrency model. In environments
where memory and CPU resources are limited, structured concurrency via
suspendable coroutines is the most efficient and robust option.

### 2025-07-13 docs/article/Sections/state-of-the-art-web-servers.tex
Post MDPI review changes.

\subsection{Web Framework Architectures and Approaches to PSSR}

@@

@@

Quarkus is a modern, cloud-native Java framework that supports both imperative
and reactive programming models. It is built on top of Vert.x,
which provides a non-blocking, event-driven runtime similar to Netty. Incoming
HTTP requests are processed asynchronously on a small, fixed-size event loop
thread pool, allowing the framework to handle many concurrent connections
efficiently without blocking threads. 
Quarkus provides support for reactive APIs like Mutiny~\cite{mutiny2021}, a
modern alternative to Spring Reactor~\cite{projectreactor}. Like Reactor,
Mutiny complies with the Reactive Streams specification~\cite{ReactiveStreams}.
For traditional imperative-style request handling, Quarkus provides integration
with JAX-RS~\cite{burke2013restful}, where resource methods can return synchronous types or
asynchronous constructs such as \texttt{CompletionStage}. To enable 
PSSR in imperative endpoints, Quarkus leverages the
JAX-RS \texttt{StreamingOutput} interface. By returning a
\texttt{StreamingOutput} instance, developers can write parts of the HTTP
response body incrementally as data becomes available. This approach allows the
server to flush HTML fragments progressively to the client, improving the
perceived responsiveness especially for long-running or large responses.
However, like other streaming mechanisms based on servlet buffers, the initial
data flush may be delayed until internal buffers (commonly around 8KB) are
filled. However, Quarkus allows configuring the buffer size to a smaller value, which
can help mitigate this issue.

>>>>

\subsection{Web Framework Architectures and Approaches to PSSR}
\label{sec:web-frameworks}

@@

The primary advantage of virtual threads lies in their ability to eliminate the
need for complex asynchronous programming models. Unlike reactive or structured
concurrency approaches that require developers to adopt complex paradigms, virtual
threads enable non-blocking execution of existing synchronous APIs without code
modification. This significantly reduces the learning curve and maintenance
burden associated with asynchronous programming while maintaining comparable
performance characteristics.

@@

Quarkus is a modern, cloud-native Java framework that supports both imperative
and reactive programming models. It is built on top of Vert.X, which provides a
non-blocking, event-driven runtime similar to Netty. Incoming HTTP requests are
processed asynchronously on a small, fixed-size event loop thread pool,
allowing the framework to handle many concurrent connections efficiently
without blocking threads. Quarkus provides support for reactive APIs like
Mutiny~\cite{mutiny2021}, a modern alternative to Spring
Reactor~\cite{projectreactor}. Like Reactor, Mutiny complies with the Reactive
Streams specification~\cite{ReactiveStreams}. For traditional imperative-style
request handling, Quarkus provides integration with
JAX-RS~\cite{burke2013restful}, where resource methods can return synchronous
types or asynchronous constructs such as \texttt{CompletionStage}. To enable
PSSR in imperative endpoints, Quarkus leverages the JAX-RS
\texttt{StreamingOutput} interface. By returning a \texttt{StreamingOutput}
instance, developers can write parts of the HTTP response body incrementally as
data becomes available. This approach allows the server to flush HTML fragments
progressively to the client, improving the perceived responsiveness especially
for long-running or large responses. Like other streaming mechanisms
based on servlet buffers, the initial data flush may be delayed until internal
buffers (commonly around 8KB) are filled. However, Quarkus allows configuring
the buffer size to a smaller value, which can help mitigate this issue.
